The Structure of Human Language Reflects Limitations of Working Memory

### 2019-04-09 paper-short.tex
revising

\title{???? The Structure of Human Language Reflects Limitations of Working Memory}
\author{Michael Hahn, Judith Degen, Richard Futrell}
\date{2019}

@@

Human languages express hierarchical thoughts into sentences, i.e., linear strings of words.
We show that the ways languages do this reflect an underlying optimization principle:
Human languages have evolved towards robustness against limitations in human working memory.

Languages differ considerably in the rules they apply to order information: English orders the object after the verb, Japanese places it before the verb.
Explaining the variation in the ways languages  order underlying hierarchical structures into linear strings of words has been one of the foci of linguistic research (CITE).
We test the hypothesis that human languages represent different solutions to the problem of efficient computation with constrained working memory.

Theoretical linguists have long suggested that the structure of human language reflects a need for efficient processing under resource limitations~\citep{berwick1984grammatical,hawkins1994performance} (also general efficiency mentioned outside functional work, e.g. \citep{chomsky2005three,hauser2002faculty}).

We directly test this hypothesis using large-scale data from 54 languages, with expert annotations of underlying hierarchical structures~\citep{nivre-universal-2017}.

We consider language understanding from the perspective of a listener processing a stream of words uttered by an interlocutor.
During comprehension, this listener will represent some amount of information about past words in their short-term memory.
We consider the question of how many bits of information this listener has to carry from one instant to the next.
We expect that a listener's ease of comprehension will be modulated by the amount of memory held about previous context:
A listener who remembers more of the past context should have less difficulty understanding, at the cost of increased working memory load.

%We use information theory to quantify the tradeoff between memory load and ease of understanding.
Experimental research has established that human processing effort on a given word is strongly predicted by its surprisal, the degree to which it cannot be predicted:
%Experimental research has firmly established Surprisal as an accurate metric of human processing effort in language comprehesion.
%The surprisal of a word in context is the degree to which it cannot be predicted:
\begin{equation}\label{eq:surp}
	-\log p(x_i|x_{1...i-1})
\end{equation}
where $x_i$ is the $i$-th word in a given utterance (or discourse).
%This quantity is known to be a strong linear predictor of human processing effort.

A listener with limited memory of prior context will not be able to compute the full conditional distribution~(\ref{eq:surp}).
We denote a listener's memory representation at time $t$ by $l_t$, which is some -- generally imperfect -- summary of the context $X_{<t}$.
The amount of information stored by the listener is the number of bits in $L_t$, i.e., its entropy $M := H[L_t]$.
This listener will experience processing effort
\begin{equation}
	S :=	-\log p(x_t|l_t)
\end{equation}
upon hearing the next word $x_t$.
%We quantify memory as the number of bits about prior input that a listener represents.

We prove a theorem relating memory to the experienced comprehension difficulty $S$.
Informally, the theorem says that languages support more efficient tradeoffs between short-term memory and comprehension difficulty when words are strongly predictive of their neighboring words.

Based on our theoretical result, we estimated the memory-surprisal tradeoff for 54 human languages, using standard estimation methods for language modeling based on neural networks~\citep{hochreiter-long-1997}.
We compared against chance by the following method.
For each language, we constructed counterfactual versions that differ from real languages in their word order, holding all other properties constant.
These counterfactual versions use randomly created grammar for linearizing hierarchical structures into sentences.
These random grammars apply consistent grammatical rules similar to human language.

Results are show in Figure~\ref{fig:results}.
Human languages lead to a more efficient tradeoff than expected by chance ($p < 0.001$).

@@

Beyond human language, our theorem applies to other stochastic processes.
To the extent that they may be affected by bounded memory in computation, we predict they should also enable efficient tradeoffs between memory and surprisal.

>>>>

\title{The Structure of Human Language Reflects Limitations of Working Memory}
\author{Michael Hahn, Judith Degen, Richard Futrell}
\date{2019}

@@

%Since short-term memory of this kind is known to be highly limited in capacity \citep{miller1956magical}, it makes sense for these capacity limits to comprise a major constraint on production and comprehension.

Human languages express hierarchical thoughts into sentences, i.e., linear strings of words.
Languages differ considerably in the rules they apply to order information: English orders the object after the verb, Japanese places it before the verb.
Explaining the variation in the ways languages use to order underlying hierarchical structures into linear strings of words has been one of the foci of linguistic research (CITE).
We show that the ways in which languages order information reflect an underlying optimization principle: Human languages have evolved towards robustness against limitations in human working memory.

%We test the hypothesis that human languages represent different solutions to the problem of efficient computation with constrained working memory.
The suggestion that the structure of human language partly reflects a need for efficient processing under resource limitations has been present in the theoretical and functional linguistics literature for decades~\citep{berwick1984grammatical,hawkins1994performance};
however, it has been hard to test this idea without ad-hoc modeling assumptions.
%Alternative?: there has been debate about what shapes human language. role played by UG/innate constraints? environment? cognition?

We directly test this hypothesis by applying a novel information-theoretic analysis to large-scale data from 54 languages, with expert annotations of underlying hierarchical structures~\citep{nivre-universal-2017}.

We consider language understanding from the perspective of a listener processing a stream of words uttered by an interlocutor.
During comprehension, this listener will represent some amount of information about past words in their short-term memory.
We consider the question of how many bits of information this listener has to carry from one instant to the next.
Human language is well known to have a large amount of context-dependence in the meanings and forms of words.
Thus, we expect that a listener's ease of comprehension will be modulated by how much memory they invest into representing information about past context:
A listener who remembers more of the past context should have less difficulty understanding, at the cost of increased working memory load.
We thus expect there to be a tradeoff between a listener's ease of language understanding and their working memory load.

%We use information theory to quantify the tradeoff between memory load and ease of understanding.
Experimental research has established that human processing effort on a given word is strongly predicted by its surprisal, the degree to which it cannot be predicted from the preceding context:
%Experimental research has firmly established Surprisal as an accurate metric of human processing effort in language comprehesion.
%The surprisal of a word in context is the degree to which it cannot be predicted:
\begin{equation}\label{eq:surp}
	-\log p(x_i|x_{1...i-1})
\end{equation}
where $x_i$ is the $i$-th word in a given utterance (or discourse).
%This quantity is known to be a strong linear predictor of human processing effort.

A listener with limited memory of prior context will not be able to compute the full conditional distribution~(\ref{eq:surp}).
We denote a listener's memory representation at time $t$ by $L_t$, which is some -- generally imperfect -- summary of the preceding context $\dots x_{t-3}x_{t-2}x_{t-1}$.
The amount of information stored by the listener is the number of bits in $L_t$, i.e., its entropy $M := H[L_t]$.
This listener will experience processing effort
\begin{equation}
	S :=	-\log p(x_t|L_t)
\end{equation}
upon hearing the next word $x_t$.
One should expect that processing effort $S$ increases for a listener who invests less bits into representing information about prior context.
%We quantify memory as the number of bits about prior input that a listener represents.
We prove a theorem describing how the shape of this tradeoff relates to language structure. %relating memory to the experienced comprehension difficulty $S$.
Informally, the theorem says that languages support more efficient tradeoffs between short-term memory and comprehension difficulty when words are strongly predictive of their neighboring words.

Based on our theoretical result, we estimated the memory-surprisal tradeoff for 54 human languages, using standard estimation methods for language modeling based on neural networks~\citep{hochreiter-long-1997}.
We compared against chance by the following method.
For each language, we constructed counterfactual versions that differ from real languages in their word order, holding all other properties constant.
These counterfactual versions use randomly created grammar for linearizing hierarchical structures into sentences.
These random grammars apply consistent grammatical rules as human language does, but have no bias towards any kind of processing efficiency or robustness.

Results are show in Figure~\ref{fig:results}.
Human languages lead to a more efficient tradeoff than expected by chance ($p < 0.0001$).
We replicated this result with different methods of estimating chance (see SI).

@@

Things to perhaps mention:

- Potentially cite \cite{PhysRevLett.115.098701} as a reference saying that physical systems have bounded memory?

- Computational efficiency has been mentioned in mainstream syntax, though it's not so clear what it means there \citep{chomsky2005three,hauser2002faculty}.

- Beyond human language, our theoretical result applies to other stochastic processes. To the extent that they may be affected by bounded memory in computation, we predict they should also enable efficient tradeoffs between memory and surprisal.
